---
title: "Projet oiseaux"
output: html_document
date: "2023-01-16"
---
```{r}
library(sf)
library(sp)
library(reshape2)
library(rgdal)
library(mapview)
library(dplyr)
library(gridExtra)
library(maptools)
library(cleangeo)
library(geosphere)
library(vegan)
library(VennDiagram)
library(ggtern)
library(terra)

```


# Data

## Bird


```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/bird')
```


We first need to extract data and associate spatial ID 
```{r}
pass <- readRDS("pass.RDS") #raw data

pass[which(pass$taxa_valid_scientific_name=='Columba livia'),] ## check if there is pigeons 

#construct space ID
spID <- as.numeric(as.factor(as.character(pass$geom$geometry)))
pass$spID <- spID #add new column of constructed id to pass data frame 

# coordinates and spID
coordSpIDAll <- st_sf(data.frame(geom = pass$geom, spID = pass$spID)) 
coordSpIDSel <- coordSpIDAll[!duplicated(coordSpIDAll$spID),]
coordSpID <- coordSpIDSel[order(coordSpIDSel$spID),]
```

We then have to select the nesting and roaming period which is from April to mid-June for the first one and from mid-July to September for the last one. 
```{r}
# Select nesting
passApril <- pass[which(pass$month_obs == 4),]
passMay <- pass[which(pass$month_obs == 5),]
passJune <- pass[which(pass$month_obs == 6 & pass$day_obs <= 15),]

passNest <- rbind(passApril, passMay, passJune)

# Select roaming
passJuly <- pass[which(pass$month_obs == 7 & pass$day_obs > 15),]
passAug <- pass[which(pass$month_obs == 8),]
passSept <- pass[which(pass$month_obs == 9),]

passRoam <- rbind(passJuly, passAug, passSept)
```

We now can build the community matrix for nesting season:
```{r}

commNestRaw <- dcast(passNest, spID ~ taxa_valid_scientific_name,
                     value.var = "value", fun.aggregate = sum)

# Convert counts to 0 - 1
commNestPA <- ifelse(commNestRaw[,-1] > 0, 1, 0)

# transform into sf object
coordNest <- coordSpID$geom[which(coordSpID$spID %in% commNestRaw$spID)]
commNest <- st_sf(data.frame(geom = coordNest, commNestPA)) 

# Check
apply(st_drop_geometry(commNest),2,unique) # Good
```

And for the roaming season:
```{r}
commRoamRaw <- dcast(passRoam, spID ~ taxa_valid_scientific_name,
                     value.var = "value", fun.aggregate = sum)

# Convert counts to 0 - 1
commRoamPA <- ifelse(commRoamRaw[,-1] > 0, 1, 0)

# transform into sf object
coordRoam <- coordSpID$geom[which(coordSpID$spID %in% commRoamRaw$spID)]
commRoam <- st_sf(data.frame(geom = coordRoam, commRoamPA)) 

# Check
apply(st_drop_geometry(commRoam),2,unique) # Good
```

Keep bird data that match between the two periods (roaming and nesting). 
```{r}
comm_nest <- commNest[,which(colnames(commNest) %in% colnames(commRoam))]
comm_roam <- commRoam[,which(colnames(commRoam) %in% colnames(commNest))]
```


We must extract birds present in green areas. We will clean green area data and do it after it's done. 

## Green areas

Data of green areas was obtained through Observatoire Grand Montreal (https://observatoire.cmm.qc.ca/produits/donnees-georeferencees/#utilisation_du_sol ) (2020). It is firstly soil occupancy data with a category of classification of "park" or "green area". 

Load the data: 
```{r}
g1<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66007-US-2020')
g2<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66023-US-2020')
g3<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66032-US-2020')
g4<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66047-US-2020')
g5<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66058-US-2020')
g6<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66062-US-2020')
g7<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66072-US-2020')
g8<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66087-US-2020')
g9<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66092-US-2020')
g10<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66097-US-2020')
g11<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66102-US-2020')
g12<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66107-US-2020')
g13<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66112-US-2020')
g14<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66117-US-2020')
g15<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66127-US-2020')
g16<-st_read('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces/raw/66142-US-2020')

```


The object g2 that doesn't have the same names as the other objects. We must renames all the columns and match it with the other dataframes. 

```{r}
colnames(g2)<-c('CODEGEOG','ZONE_CPTAQ','ETAGE_MIN','ETAGE_MAX','ETAGE_MOYE','ETAGE_MEDI',
                'ETAGE_TOTA','AIRE_MIN', 'AIRE_MAX','AIRE_MOYEN','AIRE_MEDIA','AIRE_TOTAL',
                'LOG_MIN','LOG_MAX','LOG_MOYEN','LOG_MEDIAN','LOG_TOTAL','ANNEE_MIN',
                'ANNEE_MAX','ANNEE_MOYE','ANNEE_MEDI','ANNEE_TOTA','TERRAIN_MI', 'TERRAIN_MA',
                'TERRAIN_MO','TERRAIN_ME','TERRAIN_TO','TERRM2_MIN','TERRM2_MAX','TERRM2_MOY','TERRM2_MED',
                'BAT_MIN','BAT_MAX','BAT_MOYEN','BAT_MEDIAN','BAT_TOTAL','ID','UTIL_SOL','geometry')

col_order<-match(colnames(g1),colnames(g2))
g2<-g2[,col_order]
```
Bind all data frames together to have only one for the whole island of Montreal.
```{r}
green_all<-rbind(g1,g2,g3,g4,g5,g6,g7,g8,g9,g10,g11,g12,g13,g14,g15,g16)
```

We must now extract the data for parks and greens areas (which represents in the data set the code 600 in UTIL_SOL) and golfs (1100).

```{r}
g_greens<-green_all[green_all$UTIL_SOL==600,] 
g_golf<-green_all[green_all$UTIL_SOL==1100,]

greens<-rbind(g_greens,g_golf)

```
Many green areas are not present so check with data from Données ouvertes Montréal coming from: https://donnees.montreal.ca/ville-de-montreal/grands-parcs-parcs-d-arrondissements-et-espaces-publics. We will now proceed to the same steps as with the data from observatoire Montreal.

```{r}
#open data from donn/es ouvertes montreal
setwd('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/green spaces')
green_mtl<-readRDS(file='greens')
```
We now have to extract polygons that are not present in the data from Donnees ouvertes Montreal. We first have to keep only matching columns and rename them to have the same name.
```{r}
g_obs<-greens[,c(1,39)]
g_mtl<-green_mtl[,c(1,12)]
names(g_mtl)[names(g_mtl)=='OBJECTID']<-'ID' 
```
Before doing the intersection, both dataframes must have the same projection. The intersection will tell us which polygons are the same in bith dataframes. We will then take the data from this procedure and remove it from the data of Observatoire Montreal. We will bind these two unique dataframes to obtain the map of all the green areas of Montreal. 
```{r}
g_mtl<-st_transform(g_mtl,4326)
g_obs<-st_transform(g_obs,4326)

greens_intersection<-st_intersection(g_mtl,g_obs)
green_unique<-g_obs[!(g_obs$ID %in% greens_intersection$ID.1),] # keep only one o the two IDs

green_unique<-st_transform(green_unique,4326)
greens_combine<-rbind(green_unique,g_mtl)
```
Save the outcome as an RDS file.

I now have to merge adjacent polygons since many polygons touch but are considered as different patches. I will do so with st_union, but it returns a MULTIPOLYGON, so I convert it to POLYGONS and then to sf_df.

```{r}
greens_union<-st_union(greens_combine)
greens_cast<-st_cast(greens_union,"POLYGON") 
green_areas<-st_as_sf(greens_cast)
```

I must now calculate the area of the merged polygons.
```{r}
green_areas$area<-st_area(green_areas)
class(green_areas$area) #units so I must change to numeric
```
The function st_area creates units so I must drop them in order to keep areas greater than 5 000 m2.

```{r}
green_areas$area<-as.numeric(green_areas$area)
green_areas<-green_areas[green_areas$area>5000,]
```

## Extracting birds out of green areas

We must first check if bird data and green areas have the same projection
```{r}
st_crs(comm_nest)
st_crs(comm_roam)
st_crs(green_areas)
```
It is the same projection so we don't have anything to change.

Now that I have checked the projection I can proceed to the join:
```{r}
bird_nest_green <- st_intersection(comm_nest,green_areas)
mapview(bird_nest_green)+green_areas

```
It works! Do it for roaming. 

```{r}
bird_roam_green<-st_intersection(comm_roam, green_areas)

```
We see that some areas don't have bird observations. It was decided that we would keep them for the spatial analysis, because no observations does not mean that no birds were present and could be used to "rest" when flying. 

## Tree cover 

Data of tree cover was obtained through https://observatoire.cmm.qc.ca/produits/donnees-georeferencees/#indice_canopee (2021) where tree cover is separated in high canopy (>3m) and low canopy cover (<3m). 

```{r}

tree_cover<-rast('C:/Users/greco/OneDrive - USherbrooke/Maitrise/Analyse Quantitative des données BIO6077/Projet final/data/Environmental variables/tree cover/660_IndiceCanopee_2021.tif')

```

Then we need to extract the values of pixels inside each green areas polygons. So we first have to give the same projection as the raster to the green areas and convert it to a vector.
```{r}
green_vect<-vect(green_areas_crs)
green_extract <-extract(tree_cover, green_vect)
```

We can plot it to see what it looks like and if it worked.

```{r}
plot(tree_cover)
lines(green_vect)
```
Get summary of each polygon.
```{r}
green_fac <- as.factor(green_extract [,1]) #fist column is polygons
tree_fac <- as.factor(green_extract [,2]) #second column is raster
pixels_greens<-tapply(tree_fac, green_fac, summary) #gives an array
```

Now we can put the values 3 and 4 which represents low and high canopy cover respectively in a data frame
```{r}
canopy<-data.frame()
for (i in 1:nrow(pixels_greens)) {
  canopy[i,1]<-pixels_greens[[i]][[3]] #choose value 3 for each row
  canopy[i,2]<-pixels_greens[[i]][[4]]
  colnames(canopy)<-c('low_cover','high_cover') #change column names
  print(canopy) #works
}
```


